\begin{examplebox}[label=ex:civil-service]{Example 1 --- Civil Service}
\textbf{Student writing} (Score 3):

\begin{quote}\itshape
Civil service jobs are available for all people.\,\, These jobs have fair pay from the merit system.\,\, These jobs are filled by the top applicants.\,\, This process is time consuming, but he benefits to having a civil service job out ways the steps that need to be taken.
\end{quote}

\textbf{GM rewrite} (sim $0.50$):

\begin{quote}\itshape
Civil service jobs are available for all people, offering a wide range of positions from clerks to law enforcement. These jobs are filled by the top applicants through the merit system, which ensures fair pay and fair treatment for all employees. The merit system is set up to protect the employee and provide equal and fair pay. This process is time consuming, but the benefits of having a civil service job outweigh the steps that need to be taken.
\end{quote}

\textbf{ME rewrite} (sim $0.30$):

\begin{quote}\itshape
Civil service jobs offer opportunities to men and women at various levels of government, including city, county, state, and federal. The merit system is used to recruit the most qualified individuals for each position, ensuring fair pay, training, and treatment regardless of background. With a range of job types available, from administrative roles to law enforcement and skilled trades, civil service offers secure employment options for those willing to invest the time and effort required to secure a position.
\end{quote}
\end{examplebox}
\captionof{figure}{Example of a counterfactual rewrite where Gumbel Machine (GM) remains similar in precise wording and structure to the student writing while inserting relevant intermediate sentences and clauses to improve the summary. Minimal Edit (ME) instead reorganizes the content significantly and does not preserve the student's original wording.}
\label{ex:civil-service}
