\begin{examplebox}{Example 2 --- Global Warming}
\textbf{Student writing} (Score 1):

\begin{quote}\itshape
Most people in California are more environmentally conscious than in other places in the country. They also don't believe that the government is doing enough to stop this.
\end{quote}

\textbf{GM rewrite} (sim $0.24$):

\begin{quote}\itshape
Most Californians are very concerned about global warming. They believe that the earth is warming too much and that it will affect living things on the earth. A survey was taken in California to see how people felt about global warming. The results showed that most people in California are very concerned about global warming and want the state to work on the problem. They are not happy with the federal government's handling of the issue and want the state to start working on it immediately.
\end{quote}

\textbf{ME rewrite} (sim $0.19$):

\begin{quote}\itshape
California residents are particularly concerned about the state of the planet, linking the issue to global warming. They believe that this warming poses a threat to living things. Californians want the state to take responsibility for investigating the problem.

Californians are deeply disturbed when it comes to the federal government's response to global warming. This year, they are more worried than ever, feeling the problem is getting worse. They want the state to take action immediately.

Californians want the state to start working on the problem as soon as possible, rather than waiting for the federal government to act.
\end{quote}
\end{examplebox}
\captionof{figure}{Example of a counterfactual rewrite where Gumbel Machine (GM) preserves the student's two-part key points, but then expands on the writing to provide more details. Minimal Edit (ME) instead fragments this structure across three paragraphs and introduces new temporal framing absent from the original.}
\label{ex:global-warming}
